---
title: "STAT 505 - Lesson 11 - Principal Component Analysis"
author: "Ramaa Nathan"
date: "4/3/2019"
#output: html_document
always_allow_html: yes
output: 
  github_document:
    toc: true
    toc_depth: 4
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stats)  #for prcomp
library(tidyverse) #for parsing the data
library(knitr) # for kable
library(kableExtra) # for formatting tables
library(ggplot2) #for plots
library(gridExtra) #for side by side ggplots
```
## Data Exploration
### Load File 
```{r loadfile}
places = read_table("data/places.txt", col_names=c("climate", "housing","health","crime","trans","educate","arts","recreate","econ","id") )
#str(places)
```

### Plot histograms 
```{r plothist}
ggplot(gather(places[,-10]),aes(value)) +
  geom_histogram(bins=10) +
  facet_wrap(~key, scales='free_x')
```

### Transform to normalize data 
```{r transform}
#apply log10 function - we do not need id here.. so remove it.
#places1 = places %>% select(-id) %>%
 # mutate_all(log10)
places1 <- places %>% 
  mutate_at(vars(-id),log10)
```


### Plot transformed histograms 
```{r plothist2}
ggplot(gather(places1[,-10]),aes(value)) +
  geom_histogram(bins=10) +
  facet_wrap(~key, scales='free_x')
```


```{r helper_functions}
formatAsTable <- function(df,title="Title") {
  #create the header vector
  table_header = ifelse(has_rownames(df),c(ncol(df)+1),c(ncol(df)))
  names(table_header)=c(title)
  #style the table
  df %>% kable() %>%
    kable_styling(full_width=F,bootstrap_options=c("bordered")) %>%
    add_header_above(table_header)
}

```

## Principal Component Analysis using Covariance Matrix
### Using princomp
```{r princomp}
#get the covariance matrix
places_cov <- cov(places1[,-10])
formatAsTable(data.frame(places_cov),"Covariance Matrix")

#compute PCA 
pr_cov <- princomp(places1[,-10],cor=FALSE,scores=TRUE,fix_sign=TRUE)
summary(pr_cov)

#eigen values = variance = (pr_cov$sdev)^2
# create a dataframe with eigen values, differences, variance of proportion and ciumulative proportion of variance
proportionOfVar = data_frame(EigenValue=(pr_cov$sdev)^2,
                             Difference=EigenValue-lead(EigenValue),
                             Proportion=EigenValue/sum(EigenValue),
                             Cumulative=cumsum(EigenValue)/sum(EigenValue))
formatAsTable(proportionOfVar,"Variances / Eigen Values")


#get the loadings - the Eigen Vectors
eigenVectors = pr_cov$loadings;
#create a dataframe - only for display purposes
eigenVectorsDF <- data.frame(matrix(eigenVectors,9,9))
row.names(eigenVectorsDF) <- unlist(dimnames(eigenVectors)[1])
colnames(eigenVectorsDF) <- unlist(dimnames(eigenVectors)[2])
formatAsTable(eigenVectorsDF,"Eigen Vectors")

#Display the scores of the first 10 observations
pr_cov$scores %>% head(10) %>% formatAsTable(.,"Scores")
pr_cov_combined <- cbind(places1,pr_cov$scores)


```

#### Scatter Plots of the Principal Components - using Covariance Matrix
```{r scatterplot}
#Plot the first two principal components
data.frame(pr_cov_combined) %>% 
  ggplot(mapping=aes(x=Comp.2,y=Comp.1)) +
  geom_point() +
  #geom_text(aes(label=id,hjust=0,vjust=0)) +
  labs(title = "Principal Component 1 vs 2 - using Covariance matrix") +
  theme(plot.title = element_text(hjust = 0.5))+
  xlab("Principal Component 2") +
  ylab("Principal Component 1") 
```

#### Scree Plots - using Covariance Matrix
```{r screeplot}
proportionOfVar %>% 
  ggplot(mapping=aes(x=seq(nrow(proportionOfVar)),y=EigenValue))+
  geom_point()+
  geom_line()+
  scale_x_discrete(limits=seq(nrow(proportionOfVar))) +
  xlab("Principal Component") +
  labs(title="Scree Plot - using Covariance matrix") +
  theme(plot.title = element_text(hjust = 0.5)) 
```


#### Variance Explained Plots - using Covariance Matrix
```{r proportionplot}
 proportionOfVar %>% 
  ggplot(mapping=aes(x=seq(nrow(proportionOfVar)))) + 
  geom_point(aes(y=Proportion,colour="red"))+
  geom_line(aes(y=Proportion,colour="red"))+
  geom_point(aes(y=Cumulative,colour="blue"))+
  geom_line(aes(y=Cumulative,colour="blue"),linetype="dashed")+
  scale_color_discrete(name="Type",labels=c("Cumulative","Proportion"))+ #legend
  scale_x_discrete(limits=seq(nrow(proportionOfVar))) + #xlabel
  xlab("Principal Component") +
  labs(title="Variance Explained - using Covariance Matrix") +
  theme(plot.title = element_text(hjust = 0.5)) 
```


#### Correlation between Principal Components and Variables -using Covariance Matrix
```{r}
# select only the first three principal components (so ignore column id and all columns named Comp.4 to Comp.9)
 pr_cov_combined %>% select(-matches("*.[4-9]$"),-id) %>% 
  cor() %>% as.data.frame() %>% 
  select(starts_with("Comp")) %>%
  formatAsTable(.,"Correlation between Principal Components and Variables")
```

### Using prcomp

```{r prcomp}
#compute PCA using prcomp - (SVD method)
#IMPORTANT - when using prcomp, always set center=TRUE to match the results from princomp with covaraince matrix
pr_cov2 <- prcomp(places1[,-10],scale=FALSE,center=TRUE) 

#eigen values = variance = (pr_cov$sdev)^2
# create a dataframe with eigen values, differences, variance of proportion and ciumulative proportion of variance
proportionOfVar = data_frame(EigenValue=(pr_cov2$sdev)^2,
                             Difference=EigenValue-lead(EigenValue),
                             Proportion=EigenValue/sum(EigenValue),
                             Cumulative=cumsum(EigenValue)/sum(EigenValue))
formatAsTable(proportionOfVar,"Variances / Eigen Values")


#get the laodings - the Eigen Vectors
eigenVectors = pr_cov2$rotation;
#create a dataframe - only for display purposes
eigenVectorsDF <- data.frame(matrix(eigenVectors,9,9))
row.names(eigenVectorsDF) <- unlist(dimnames(eigenVectors)[1])
colnames(eigenVectorsDF) <- unlist(dimnames(eigenVectors)[2])
formatAsTable(eigenVectorsDF,"Eigen Vectors")

#Display the scores of the first 10 observations
pr_cov2$x %>% head(10) %>% formatAsTable(.,"Scores")

```


## Principal Component Analysis using Correlation Matrix - Standardized values 
```{r princompCorr}
#compute PCA 
pr_corr <- princomp(places1[,-10],cor=TRUE,scores=TRUE,fix_sign=TRUE)
summary(pr_corr)

#eigen values = variance = (pr_corr$sdev)^2
# create a dataframe with eigen values, differences, variance of proportion and ciumulative proportion of variance
proportionOfVar = data_frame(EigenValue=(pr_corr$sdev)^2,
                             Difference=EigenValue-lead(EigenValue),
                             Proportion=EigenValue/sum(EigenValue),
                             Cumulative=cumsum(EigenValue)/sum(EigenValue))
formatAsTable(proportionOfVar,"Variances / Eigen Values")


#get the laodings - the Eigen Vectors
eigenVectors = pr_corr$loadings;
#create a dataframe - only for display purposes
eigenVectorsDF <- data.frame(matrix(eigenVectors,9,9))
row.names(eigenVectorsDF) <- unlist(dimnames(eigenVectors)[1])
colnames(eigenVectorsDF) <- unlist(dimnames(eigenVectors)[2])
formatAsTable(eigenVectorsDF,"Eigen Vectors")

#Display the scores of the first 10 observations
pr_corr$scores %>% head(10) %>% formatAsTable(.,"Scores")
pr_corr_combined <- cbind(places1,pr_corr$scores)
```

#### Scatter Plots of the Principal Components - using Correlation Matrix
```{r scatterplotCorr}
#Plot the first two principal components
data.frame(pr_corr_combined) %>% 
  ggplot(mapping=aes(x=Comp.2,y=Comp.1)) +
  geom_point() +
  #geom_text(aes(label=id,hjust=0,vjust=0)) +
  labs(title = "Principal Component 1 vs 2 - using Correlation matrix") +
  theme(plot.title = element_text(hjust = 0.5))+
  xlab("Principal Component 2") +
  ylab("Principal Component 1") 
```

#### Scree Plots - using Correlation Matrix
```{r screeplotCorr}
proportionOfVar %>% 
  ggplot(mapping=aes(x=seq(nrow(proportionOfVar)),y=EigenValue))+
  geom_point()+
  geom_line()+
  scale_x_discrete(limits=seq(nrow(proportionOfVar))) +
  xlab("Principal Component") +
  labs(title="Scree Plot - using Correlation matrix") +
  theme(plot.title = element_text(hjust = 0.5)) 
```


#### Variance Explained Plots -using Correlation Matrix
```{r proportionplotCorr}
 proportionOfVar %>% 
  ggplot(mapping=aes(x=seq(nrow(proportionOfVar)))) + 
  geom_point(aes(y=Proportion,colour="red"))+
  geom_line(aes(y=Proportion,colour="red"))+
  geom_point(aes(y=Cumulative,colour="blue"))+
  geom_line(aes(y=Cumulative,colour="blue"),linetype="dashed")+
  scale_color_discrete(name="Type",labels=c("Cumulative","Proportion"))+ #legend
  scale_x_discrete(limits=seq(nrow(proportionOfVar))) + #xlabel
  xlab("Principal Component") +
  labs(title="Variance Explained - using Correlation Matrix") +
  theme(plot.title = element_text(hjust = 0.5)) 

```

#### Correlation between Principal Components and Variables - using Correlation Matrix
```{r corrCorr}
# select only the first three principal components (so ignore column id and all columns named Comp.4 to Comp.9)
 pr_corr_combined %>% select(-matches("*.[4-9]$"),-id) %>% 
  cor() %>% as.data.frame() %>% 
  select(starts_with("Comp")) %>%
  formatAsTable(.,"Correlation between Principal Components and Variables")
```
