---
title: "STAT505 Hierarchical Cluster "
author: "Ramaa Nathan"
date: "4/20/2019"
#output: html_document
always_allow_html: yes
output:
  github_document:
    toc: true
    toc_depth: 4
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stats)  
library(tidyverse) #for parsing the data
library(knitr) # for kable
library(kableExtra) # for formatting tables
library(ggplot2) #for plots
library(gridExtra) #for side by side ggplots
library(ggdendro) #for plotting dendrograms
```

This is a R impelmentation of the hierarchical clustering example used in the on-line notes of STAT 505.
This example uses ecological data from Woodyard Hammock, a beech-magnolia forest in northern Florida. The data involve counts of the numbers of trees of each species in n = 72 sites. A total of 31 species were identified and counted, however, only p = 13 of the most common species were retained and are listed below. They are:

Species Code | Latin Name | Common Name
------|-------|--------
carcar  |	Carpinus caroliniana  | Ironwood 
corflo |	Cornus florida |	Dogwood
faggra |	Fagus grandifolia |	Beech
ileopa |	Ilex opaca |	Holly
liqsty |	Liquidambar styraciflua	| Sweetgum
maggra |	Magnolia grandiflora |	Magnolia
nyssyl |	Nyssa sylvatica	| Blackgum
ostvir |	Ostrya virginiana |	Blue Beech
oxyarb |	Oxydendrum arboreum |	Sourwood
pingla |	Pinus glabra	| Spruce Pine
quenig |Quercus nigra	 |Water Oak
quemic |	Quercus michauxii	| Swamp Chestnut Oak
symtin |	Symplocus tinctoria	|Horse Sugar


The objctive of this data is to  group sample sites together into clusters  that share similar species compositions as determined by some measure of association. In this document, we will be using agglomerative heracrchical clustering techniques. 


## Data Exploration
### Load File 
```{r loadfile}
wood = read_table("data/wood.txt", col_names=c("x", "y","acerub","carcar","carcor","cargla","cercan","corflo","faggra","frapen",
  "ileopa", "liqsty","lirtul","maggra","magvir","morrub","nyssyl","osmame","ostvir","oxyarb","pingla","pintae","pruser", "quealb","quehem","quenig",
  "quemic","queshu","quevir","symtin","ulmala","araspi","cyrrac") )

#Dimensions of the table read in
dim(wood)

#We are only interested in 13 species (here they will be considered as 13 variables)
excluded_vars <- c("acerub","carcor","cargla","cercan","frapen","lirtul","magvir","morrub","osmame","pintae","pruser",
                   "quealb","quehem","queshu","quevir","ulmala","araspi","cyrrac")

#Filter out the variables
wood <- wood %>% 
  mutate(ident = row_number()) %>%
  select (-one_of(excluded_vars))

#Check the dimensions and column names of the final dataframe
dim(wood)
colnames(wood)
```



```{r helper_functions, echo=FALSE}
#Helper Functions
formatAsTable <- function(df,title="Title") {
  #create the header vector
  table_header = ifelse(has_rownames(df),c(ncol(df)+1),c(ncol(df)))
  names(table_header)=c(title)
  #style the table
  df %>% kable() %>%
    kable_styling(full_width=F,bootstrap_options=c("bordered")) %>%
    add_header_above(table_header)
}
```

## Hierarchical Clustering 
### Complete Linkage 
We will use the Euclidean distance as the dissimilarity measure. In R, the dist(<matrix>) fucntion computes the Euclidean distance between the rows of the matrix. In our example, we only want to compute the distances for the 13 variables. So, we will use wood[,3:15]
```{r}
# we only want to use the columns 3 throught 15 for computing the disimilarities
hc_complete <- hclust(dist(wood[,3:15]),method="complete")

#extract the dendrogram 
hc_complete_den <- as.dendrogram(hc_complete)

# use ggdendrogram (basaed on ggplot2) to plot the dendrograms
ggdendrogram(hc_complete,size=2,rotate=TRUE,scale=2) +
  labs(title="Dendrogram - Complete Linkage") +
  xlab("Height") + ylab("Sites")
```

We want to next find the history of the different merges in forming the different clusters. From the documentation, the order of merges is as follows: Row i of merge describes the merging of clusters at step i of the clustering. If an element j in the row is negative, then observation -j was merged at this stage. If j is positive then the merge was with the cluster formed at the (earlier) stage j of the algorithm. Thus negative entries in merge indicate agglomerations of singletons, and positive entries indicate agglomerations of non-singletons. The Max distance is the maximum distance computed between the two clusters as required for complete linkage

```{r}
hc_complete_hist <- as_data_frame(cbind(hc_complete$merge,hc_complete$height))
colnames(hc_complete_hist)<- c("ClusterA","ClusterB","Max Distance")
hc_complete_hist <- hc_complete_hist %>%
  mutate ("Merge Order" = row_number()) %>% 
  select(`Merge Order`, ClusterA, ClusterB, `Max Distance`)
formatAsTable(hc_complete_hist,"Clustering History")
```

Cluster numbers
```{r}
hc_complete_clusters = as_data_frame(cbind(wood$ident,cutree(hc_complete, k=6)))
colnames(hc_complete_clusters) <- c("Identity","Cluster")
formatAsTable(hc_complete_clusters,"Clusters")
```

Interpret the Clusters by doing an ANOVA on all the variables

```{r}
wood_cluster <- wood %>% mutate(cluster = cutree(hc_complete,k=6))

# WE could do an ANOVA fit with one model statement - but it is difficult to extract the individual p-values.
manova_fit <- manova(cbind(carcar, corflo, faggra, ileopa, liqsty, maggra, nyssyl, ostvir, oxyarb, 
     pingla,quenig, quemic, symtin) ~ factor(cluster), data=wood_cluster)
str(summary.aov(manova_fit))

# So, do a separate anova for each response variable and create a list of the fits
fitlist = sapply(wood_cluster[,3:15],function(resp) summary(aov(resp ~ factor(cluster), data=wood_cluster)))
response_names <- names(fitlist)
str(fitlist[[1]]["Pr(>F)"])
sig_responses = NULL
for(l in seq_along(fitlist)) {
  #print(fitlist[l])
  if(fitlist[[l]]$"Pr(>F)"[1] < (0.05/13)) {
    print(paste(response_names[l],": ",fitlist[[l]]$"Pr(>F)"[1]))
    sig_responses = append(sig_responses,response_names[l])
  }
}
#sig_responses
paste("Significant Variables:")
sig_responses


```

Get the cluster means of only the significant response variables
```{r}
cluster_means <- wood_cluster %>% group_by(cluster) %>% summarise_at(vars(sig_responses),mean)
formatAsTable(cluster_means,"Cluster Means of Significant Variables")
```




